Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm

### 2014-08-19 paint-contour-tree.tex
Changes to 5.3

\newcommand{\XSays}[2]{{
      {$\rule[-0.12cm]{0.2in}{0.5cm}$\fbox{\tt
            #1:} }
      \textcolor{red}{#2}
      \marginpar{\textcolor{blue}{#1}}
      {$\rule[0.1cm]{0.3in}{0.1cm}$\fbox{\tt
            end}$\rule[0.1cm]{0.3in}{0.1cm}$}
      }
   }
\newcommand{\Ben}[1]{{\XSays{Ben}{#1}}}

@@

\newcommand{\pathTree}{P_{\mathcal{T}}}
\begin{definition}
Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally defines a tree\footnote{Please excuse the 
overloading of the term 'tree', it is the most natural term to use here.} of there own.  Specifically, in the original 
tree $T$ contract each path down to its root.  Call the resulting tree the \emph{path tree} of $T$ corresponding to 
the path decomposition $P(T)$, and denote it by $\pathTree(T)$.

@@

\begin{definition}
 Let $p$ be a path in a path decomposition $P(T)$.  Call a light path in $P(T)$ reachable from $p$ if there is a path in the tree
 $T$ from some vertex in $p$ which only moves to vertices of larger height and only vertices which lie in light paths.
 Call this set of reachable paths the \emph{light path set} of $p$.  
 Consider the natural tree over the paths in the light path set of $p$, where each vertex in the tree is a light path 
 $p$ is the root.  We call this tree the \emph{light path tree} of $p$
\end{definition}

Let $\mathcal{L}$ be a light path tree of some light path $p$ in a maximum path decomposition $\mathcal{P}(T)$.
For a path $v\in \mathcal{L}$, let $H_v$ denote the heap at the root of $v$, and let $H_{v_f}$ denote the heap 
at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of light (resp. heavy) paths adjacent to $v$ in $T$
(i.e. $S_l(v)$ are the children of $v$ in $\mathcal{L}$).  Then define the cost of $v$ as 
$c(v) = H_{v_f}+\sum_{i\in S_h(v)} H_i$.

We now define a weighted version of $\mathcal{L}$ based on these cost.  First we augment $\mathcal{L}$ by attaching one 
leaf to each vertex in $\mathcal{L}$.  Now for each newly created leaf (which are the only leaves now) assign the cost 
$c(v)$ defined above, where $v$ is that leaf's corresponding parent.  Now for any vertex $u$ in this tree, define 
its weight to be the sum of the weights of its children (where at the leaves the weights are just the costs).  
To simplify notation, we will also refer to this new weighted tree as $\mathcal{L}$.

As the weight $w(v)$ includes the sizes of the heaps of all leaves in the subtree rooted at $v$, we have the following.

\begin{claim}
For any $v\in \mathcal{L}$, $w(v)\geq |H_v|$
\end{claim}

Let $W= \sum_{v\in \mathcal{L}} w(v)$.  Suppose one could prove $W=O(H_r)$, where $r$ is the root of $\mathcal{L}$.  
Then by the above claim, 
$\sum_{v\in \mathcal{L}} H_v = O(H_r)$, and so by convexity and \Cor{heavyLight}, we could argue the entire cost 
of all light paths in the light path tree can be charge to a heavy path.  So proving $W=O(H_r)$ is our new goal.

>>>>

\newcommand{\XSays}[2]{{
      {$\rule[-0.12cm]{0.2in}{0.5cm}$\fbox{\tt
            #1:} }
      \textcolor{red}{#2}
      \marginpar{\textcolor{blue}{#1}}
      {$\rule[0.1cm]{0.3in}{0.1cm}$\fbox{\tt
            end}$\rule[0.1cm]{0.3in}{0.1cm}$}
      }
   }
\newcommand{\Ben}[1]{{\XSays{Ben}{#1}}}
\newcommand{\pth}[2][\!]{#1\left({#2}\right)}

@@

\newcommand{\pathTree}{P_{\mathcal{T}}}
\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{T}}}
\begin{definition}
Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally defines a tree\footnote{Please excuse the 
overloading of the term 'tree', it is the most natural term to use here.} of there own.  Specifically, in the original 
tree $T$ contract each path down to its root.  Call the resulting tree the \emph{path tree} of $T$ corresponding to 
the path decomposition $P(T)$, and denote it by $\pathTree(T)$.

@@

We now prove that in a maximum path decomposition, when going between adjacent light paths, the heap size decreases by a constant fraction.

\begin{lemma}
\label{lem:geometric}
 Let $p$ and $q$ be two light paths such that $p$ is the parent of $q$ in the path tree $\pathTreeA$.  
 Then $|H_q|\leq \frac{3}{5} |H_p|$, where $H_p$ and $H_q$ are the heaps at the roots of $p$ and $q$, respectively. 
\end{lemma}
\begin{proof}
 Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
 i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
 As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
 the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
 Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
 Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
 Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
\end{proof}

Naturally, this geometric series behavior can be extended to more than two adjacent light paths, thus motivating 
the following definition.

\begin{definition}
 Let $p$ be a path in a path decomposition $P(T)$, whose corresponding path tree is $\pathTree$.
 Call a light path $q\in P(T)$ light reachable from $p$ if the path from $q$ to $p$ in $\pathTree$ 
 contains only vertices which correspond to light paths.  Call the subtree of $\pathTree$ consisting 
 of all light reachable paths from $p$ the \emph{light path tree} of $p$.
\end{definition}

Let $\mathcal{L}$ be a light path tree of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
Our goal is to use \Lem{geometric}  to show that $\sum_{v\in \mathcal{L}} |H_v| = O(|H_r|)$, namely the total sum of the sizes of the heaps 
at the roots of all paths in the light path tree is on the order of the size of the heap at the root of $r$.  
ADD THE ARGUMENT saying how this fact will be used.

For a path $v\in \mathcal{L}$, let $H_v$ denote the heap at the root of $v$, and let $H_{v_f}$ denote the heap 
at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of light (resp. heavy) paths adjacent to $v$ in $\mathcal{L}$.

We want to understand how the heap size changes as we move around the paths in $\mathcal{L}$.  Specifically, there are 
three places the vertices in the heap along a light path can come from; adjacent light paths, adjacent heavy paths, and the initial heap at the leaf of the path.  
Given this observation we now define a weight function which upper bounds the size of the heap at the root of any path in $\mathcal{L}$.

First, for all $v\in \mathcal{L}$ we define the local cost function $c(v) = H_{v_f}+\sum_{i\in S_h(v)} H_i$, this is the addition to the heap at $v$ which 
was not already charged to descendant light paths.  We now define a weighted version of $\mathcal{L}$ based on these costs.  
To do so, we first we augment $\mathcal{L}$ by attaching one 
leaf to each vertex in $\mathcal{L}$.  Now for each newly created leaf (which are the only leaves now) assign the cost 
$c(v)$ defined above, where $v$ is that leaf's corresponding parent.  
Now we can define a weight function $w(u)$ over any node $u$ in the tree, which is defined to be the total cost of all leaves in its subtree.   
To simplify notation, we now use $\mathcal{L}$ to refer to this new weighted tree (rather than its unweighted counterpart).

As the weight $w(v)$ includes the sizes of the heaps of all contributing heavy paths and leaves in the subtree rooted at $v$, we have the following.

\begin{claim}
For any $v\in \mathcal{L}$, $w(v)\geq |H_v|$.
\end{claim}

Let $W= \sum_{v\in \mathcal{L}} w(v) \geq \sum_{v\in \mathcal{L}} |H_v|$.  Suppose we could show that conversely $W=O(H_r)$.  
Then we would have $\sum_{v\in \mathcal{L}} H_v = O(H_r)$, which was our original goal.  So proving $W=O(H_r)$ is now our new goal.

To this end, first observe that $|H_r|\geq w(r)-\sum_{v\in \mathcal{L}} |v|$.  This holds by the same argument as in \Lem{adjacent}, namely, 
$w(r)$ by definition counts all the contributing heavy path heaps and initial leaf heaps that contribute to the heap at $r$ and $\sum_{v\in \mathcal{L}} |v|$
is an upper bound on the number of vertices that could have been removed from one of these heaps before reach the root of $r$ (since a vertex in a heap can only 
get removed if it appears on one of the paths).

As all paths in $\mathcal{L}$ are light, we thus have the following.

\begin{claim}
$|H_r|\geq w(r)-\sum_{v\in \mathcal{L}} \sqrt{H_v}/100 \geq w(r)-\sum_{v\in \mathcal{L}} \sqrt{w(v)}/100$
\end{claim}

We will need the following lemma.

\begin{lemma}
\label{lem:distribution}
Let $C\subseteq \mathcal{P}$ be the subset of paths that are leaves in $\mathcal{L}$.

Then $\sum_{v\in \mathcal{P}} \sqrt{w(v)}/100 \leq \sum_{c\in C} \sqrt{w(c)}/20$
\end{lemma}

\begin{proof}
The idea is to recursively redistribute the contribution to the sum of each internal vertex to its children, until all the weight sits on the children.
Specifically, let $v$ be some path in $\mathcal{L}$.  Its contribution to the sum is $\sqrt{w(v)}/100$, which we will call the value at $v$.  
We redistribute this value to its children, 
in proportion by the weight of each child.  For example, if $v$ has a child $x$ then the fraction it gets assigned is $w(x)/w(v)$ 
(recall that by definition $w(v)$ is the sum of the weights of its children and so the total value will get reassigned to the children).

Suppose we keep redistributing the value of $v$ until it reaches the leaves of $\mathcal{L}$.  Let's see how much gets assigned to a given leaf.
Specifically, let $u_0, u_1, \ldots, u_k$ be a path in $\mathcal{L}$ (of paths from the set $\mathcal{P}$), such that $v=u_0$,
$u_{i+1}$ is child of $u_i$ for all $i$, and $u_k$ is a leaf of $\mathcal{L}$.

Then the total value from $u_0$ that ends up at $u_k$ is 
\[
\frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_1)}{w(u_0)} \frac{w(u_2)}{w(u_1)} \cdots \frac{w(u_k)}{w(u_{k-1})} }
= \frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_k)}{w(u_0)} } = \frac{w(u_k)}{100\sqrt{w(u_0)}}
\]

For a given leaf in $\mathcal{L}$, we can calculate the total value reassigned to it from it its ancestors.
Specifically, let $u_0, u_1, u_2, \dots, u_m$ be a root to leaf path in $\mathcal{L}$ (i.e. $u_0$ is the root $r$ of $\mathcal{L}$).

Then by the above calculation and \Lem{geometric}, the total reassigned to $u_m$ is

\Ben{There is an issue here.  \Lem{geometric} is in terms of the heaps not the weights.}

\[
 \sum_{i=0}^m \frac{w(u_m)}{100\sqrt{w(u_i)}} \leq \frac{w(u_m)}{100} \sum_{i=0}^m 1 \Big/ \sqrt{\pth{\frac{5}{3}}^iw(u_m)} 
 \leq \frac{\sqrt{w(u_m)}}{100} \sum_{i=0}^{\infty} \pth{\frac{3}{5}}^{i/2} \leq \frac{\sqrt{w(u_m)}}{20}
\]

The claim now follows by summing this quantity over the children of $\mathcal{L}$.
\end{proof}

By definition, $w(r) = \sum_{c\in C} w(c)$, where $C$ is the subset of leaves of $\mathcal{L}$.  
Therefore combining the above lemma and claim gives the following.

\begin{claim}
 $|H_r|\geq w(r)/2$
\end{claim}

\begin{lemma}
 $W=O(H_r)$
\end{lemma}
\begin{proof}
MISSING.  Is this what we are claiming?  I don't see the proof.
\end{proof}
%
%
% So $w(r)$ upper bound $H_r$, but by how much?  Consider a path $v\in \mathcal{L}$.  
% As we move from the leaf to the root of $v$ we now want to understand 
% how the heap can decreases.  Now the only way a vertex in the heap can leave that 
% heap is if that vertex is one of the vertices we traverse in $v$, i.e. the total 
% loss in vertices from the heap along $v$ is bounded by $|v|$.  
% Therefore, $|H_r|\geq W-\sum_{v\in \mathcal{L}} |v|$.  
% (This is an extension of \Lem{adjacent}, and probably should be merged with it.)
% As $\mathcal{L}$ consists of only light paths we have $|v|\leq \sqrt{|H_v|}/100 \leq \sqrt{w(v)}/100$, 
% and therefore $|H_r|\geq W-\sum_{v\in \mathcal{L}} \sqrt{w(v)}/100$.
